\pkg{sparseHessianFD}: An R Package for Estimating Sparse Hessian Matrices

### 2016-12-23 nobuild/jss_revision/sparseHessianFD.Rnw
still working on jss revision

The \pkg{sparseHessianFD} package is a tool for estimating
sparse Hessians using finite differences of gradients.  Section \ref{sec:numdiff} will
cover the specifics, but the basic idea is as follows.  Consider a
function $f(x)$, its gradient $\nabla{f(x)}$, and its Hessian
$\hess{}{f(x)}$, for $x\in\mathbb{R}^M$.  Define the derivative vector
as the transpose of the gradient, and a vector of partial derivatives, so
${\parD{}{}{f(x)}=\nabla f(x)^\top=\left(\parD{}{1}{},\mathellipsis,\parD{}{M}{}\right)}$.   (Throughout the paper, we will try
to reduce notational clutter by referring to the derivative and
Hessian as $\parD{}{}{}$ and $\hess{}{}$, respectively, without the
$f(x)$ symbol).  Let $e_m$ be a vector of zeros, except with a 1 in the $m$th
element, and let $\delta$ be a sufficiently small scalar
constant. A linear approximation to the $m$th column of the Hessian is ${\hess{m}{}\approx\left(\nabla f(x+\delta
  e_m)- \nabla f(x)\right)/\delta}$.  Estimating a dense Hessian in this way involves at least $M+1$ calculations of the
gradient: one for the gradient at $x$, and one after perturbing each
of the $M$ elements of $x$, one at a time.\footnote{More accurate
  approximations using finite differences require more gradient evaluations.  We will consider
  only forward finite differences. In Section \ref{sec:complex} we
  introduce the complex step method that, under certain conditions,
  can generate more accurate estimates with the same number of
  gradient evaluations using forward finite differences.} In a sparse matrix, most of the
elements are constrained to zero.  Depending
on the sparsity pattern of the Hessian, those constraints may let us recover
the Hessian with fewer gradient evaluations by perturbing multiple
elements of $x$ together.  For some sparsity patterns, estimating a Hessian in
this way can be profoundly efficient.  In fact, for the hierarchical models that we consider in
this paper, the number of gradient evaluations does not increase with
the number of additional heterogeneous units.

@@

This article proceeds as follows.  First, we present
some background information about numerical differentiation, and
sparse matrices in \proglang{R}, in Section~\ref{sec:background}.  In
Section ~\ref{sec:using}, we explain how to use the package.
Section~\ref{sec:algorithms} explains the underlying algorithms, and
Section~\ref{sec:timing} demonstrates the scalability of those algorithms.

@@

The \pkg{sparseHessianFD} algorithms assign variables to groups
before computing the values of the Hessian.  This is why the sparsity
pattern needs to be provided in advance.  If a non-zero element is
omitted from the sparsity pattern, the resulting estimate of the
Hessian will be incorrect.  The only problems with erroneously including a zero
element in the sparsity pattern are a possible lack of efficiency
(e.g., an increase in the number of gradient evaluations), and that
the estimated value might be close to, but not exactly, zero. The
algorithms for assigning decision variables to groups, and for
extracting nonzero Hessian elements via substitution, are described in
Section \ref{sec:algorithms}.

@@

From the log posterior density in Equation \ref{eq:LPD}, we can see
that $\parD{2}{\beta_i, \beta_i}\neq 0$ (one element of $\beta_i$ could be
correlated with another element of $\beta_i$), and that, for all $i$,
$\parD{2}{\beta_i, \mu}\neq 0$ (because $\mu$ is the prior mean of each
$\beta_i$).  However, since the $\beta_i$ and $\beta_j$ are independently distributed, and the $y_i$ are
conditionally independent, the cross-partial derivatives
$\parD{2}{\beta_i, \beta_j}=0$ for all $i\neq j$. When $N$ is much
greater than $k$, there will be many more zero cross-partial
derivatives than non-zero.  Regardless of how each $\parD{2}{}$ is
mapped to $\hess{}{}$ through the indexing function, the Hessian of the log
posterior density will be sparse.

@@

{\setlength{\extrarowheight}{.25em}
\begin{table}[ptb]
\begin{tabularx}{\linewidth}{>{\bfseries}r X}
  \toprule
x& A numeric vector, with length $M$ at which the object will be
  initialized and tested.\\
fn,gr& \proglang{R} Functions that return the value of the
  objective function, and its gradient. The first argument is the numeric
  variable vector.  Other named arguments can be passed to \func{fn}
  and {gr} as well (see the \funcarg{...} argument below).\\
rows, cols& Sparsity pattern: integer vectors of the row and column indices of
  the non-zero elements in the \emph{lower triangle} of the Hessian.\\
direct& This argument is deprecated, and is included only for
  backwards compatibility with earlier versions.\\
delta& The perturbation amount for finite differencing of the
  gradient to compute the Hessian (the $\delta$ in Section \ref{sec:numdiff}).  Defaults to
  \code{sqrt(.Machine\$double.eps)}.\\
index1& If \variable{TRUE} (the default), \funcarg{row} and \funcarg{col} use one-based
  indexing.  If \variable{FALSE}, zero-based indexing is used.\\
  complex&If \variable{TRUE}, the complex step method is used.
                If \variable{FALSE} (the default), a simple finite
                differencing of gradients is used.\\
...& Additional arguments to be passed to \func{fn} and \func{gr}.\\
 \bottomrule
\end{tabularx}
\caption{Arguments to the sparseHessianFD initializer.}\label{tab:init}
\end{table}
}

@@

Now we can estimate the Hessian for the
log posterior density of the model from Section \ref{sec:example}.
For demonstration purposes, \pkg{sparseHessianFD} includes functions that compute the value
(\func{binary.f}), the gradient (\func{binary.grad}) and the Hessian
(\func{binary.hess}) of this model. We will treat the result from \func{binary.hess} as a
  ``true'' value against which we will compare the numerical
  estimates.

@@

The \pkg{sparseHessianFD} package also allows for
estimation of the Hessian using the complex step method
\citep{SquireTrapp1998} as an alternative to finite differencing. In
this method, the definition of $\Mat{Y}_c$ in Equation \ref{eq:Yg} is replaced with
\begin{align}
  \label{eq:complexYg}
  \Mat{Y}_c&=\Imag(\nabla f(x+iG_c))
\end{align}
where $i=\sqrt{-1}$ and $\Imag$ returns the imaginary part of a complex
number. The advantage of this method is that there is no longer a
subtraction operation to generate truncation error. Thus, $\delta$ can
be made arbitrarily small, and the resulting estimate of the Hessian
is more accurate.

To use the complex step method, include the argument
\funcarg{complex=TRUE} in the call to the initializer of the \class{sparseHessianFD}
object. The functions passed as \funcarg{fn} and \funcarg{gr} must both
accept a complex argument, and return a complex
result, even though we are differentiating a real-valued function. Although \pkg{base} \proglang{R} supports complex arguments
for basic arithmetic, trigonometric, logarithmic and exponential
functions, many other common functions (e.g.,  \func{gamma}, \func{log1p},
\func{expm1}, and the probability distribution functions) do not have
complex implementations. Furthermore, the complex step method is valid
only if the function is holomorphic
(complex differentiable). The methods in \class{sparseHessianFD} do
\emph{not} check that this is the case for the function at hand. We convey the
following warning from the documentation of the \pkg{numDeriv} package
\citep{R_numDeriv}, which also implements the complex step method:
``Avoid this method if you do not know that your function is
suitable. Your mistake may not be caught and the results will be
spurious.''

The log posterior density in Equation \ref{eq:LPD} is holomorphic, so we
can estimate its Hessian using the complex step method, and compute
the mean relative difference from the true Hessian.

@@

As far as we know, \pkg{numDeriv} \citep{R_numDeriv} is the only other \proglang{R} package
that computes numerical approximations to derivatives.  Like
\pkg{sparseHessianFD} it includes functions to compute Hessians from
user-supplied gradients (through the \func{jacobian} function), and
implements both the finite differencing and complex step methods.  Its
most important distinction from \pkg{sparseHessianFD} is that it treats all Hessians as dense.  Thus, we
will use  \pkg{numDeriv} as the baseline against which we can compare the performance of
\pkg{sparseHessianFD}.  All of the subsequent speed and scalability
results were generated on a 2013-vintage Apple Mac Pro with a 2.7 GHz
12-core Intel Xeon E5 processor and 64 GB of RAM.

To prepare Table~\ref{tab:numDeriv}, we
estimated Hessians of the log posterior density in Equation~\ref{eq:LPD} with different numbers of heterogeneous
units ($N$) and within-unit parameters ($k$).  As in the previous
section, the total number of
variables is $M=(N+1)k$.  Table~\ref{tab:numDeriv} shows the mean
and standard deviations (across 20 replications) for the time (in
milliseconds) to compute a Hessian using each package.  The difference in run times is dramatic,
especially because the computation time for \pkg{numDeriv} grows
quadratically in the number of variables.  The setup time for
\pkg{sparseHessianFD} was about 7 milliseconds for all cases.  Times
were collected using the \pkg{microbenchmark} package \citep{R_microbenchmark}.

\begin{table}[ptb]\centering
  \begin{tabularx}{1\linewidth}{@{\extracolsep{\fill} }cccp{2pt}rrrrp{2pt}rrrr}
  \toprule
&&&&\multicolumn{4}{c}{\bfseries{Hessian
    time}}&&\multicolumn{4}{c}{\bfseries{Hessian/Gradient ratio}}\\
&&&&\multicolumn{2}{c}{\pkg{numDeriv}}&\multicolumn{2}{c}{\pkg{sparseHessianFD}}&&\multicolumn{2}{c}{\pkg{numDeriv}}&\multicolumn{2}{c}{\pkg{sparseHessianFD}}\\
$N$&$k$&$M$&&mean&sd&mean&sd&&mean&sd&mean&sd\\
  \cmidrule{1-3}\cmidrule{5-8}\cmidrule{10-13}
   9 &    2 &   20 &&  363 & 6.4 &  2.8 & 2.2 &&  871 & 63 & 6.5 & 4.2 \\
    12 &    2 &   26 &&  629 &  8.9 &  2.9 & 2.1 && 1469 &  124 & 6.5 & 3.3 \\
     9 &    3 &   30 &&  819 & 27.7 &  3.6 & 2.2 && 1974 &  161 & 8.4 & 3.7 \\
    15 &    2 &   32 &&  973 & 25.8 &    3.0 & 2.2 && 2191 &  209 &  6.5 & 3.4 \\
    12 &    3 &   39 && 1407 & 39.3 & 3.5 & 2.0 && 3291 &  285 &    8.0 & 3.5 \\
     9 &    4 &   40 && 1445 & 28.8 & 4.3 & 2.1 && 3453 &  292 & 9.9 & 3.1 \\
    15 &    3 &   48 && 2138 & 92.7 & 3.5 & 2.0 && 4861 &  341 & 7.8 & 3.4 \\
    12 &    4 &   52 && 2400 & 98.6 & 4.3 & 2.1 && 5679 &  503 & 10.1 & 3.7 \\
    15 &    4 &   64 && 3643 &  209.0 & 4.2 & 2.2 && 8945 &  828 & 10.0 &  2.8 \\
   \bottomrule
\end{tabularx}
\caption{Computation times (milliseconds) for computing Hessians using the
  \pkg{numDeriv} and \pkg{sparseHessianFD} packages, across 200 replications. Rows are ordered by the
  number of variables.}\label{tab:numDeriv}
\end{table}

@@

Computation times
for the function and gradient, as well as the setup and
partitioning times for the \class{sparseHessianFD} object, grow
linearly with the number of heterogenous units.  The time
for the Hessian grows linearly as well, and that might be partially
surprising.  We saw in Section \ref{sec:coloring} that adding
additional heterogeneous units in a hierarchical model does not
increase the number of required gradient evaluations.  So we might
think that the time
to compute a Hessian should not increase with $N$ \emph{at all}.  The
reason it does is that each gradient evaluation takes longer.  The
plot of the ratio of Hessian-to-gradient run times is relatively flat as the size
of the dataset increases.  We can conclude that the
\pkg{sparseHessianFD} algorithms are quite efficient and scalable for
hierarchical models.

>>>>

The \pkg{sparseHessianFD} package is a tool for estimating
sparse Hessians using finite differences of gradients.  Section~\ref{sec:numdiff} will
cover the specifics, but the basic idea is as follows.  Consider a
function $f(x)$, its gradient $\nabla{f(x)}$, and its Hessian
$\hess{}{f(x)}$, for $x\in\mathbb{R}^M$.  Define the derivative vector
as the transpose of the gradient, and a vector of partial derivatives, so
${\parD{}{}{f(x)}=\nabla f(x)^\top=\left(\parD{}{1}{},\mathellipsis,\parD{}{M}{}\right)}$.   (Throughout the paper, we will try
to reduce notational clutter by referring to the derivative and
Hessian as $\parD{}{}{}$ and $\hess{}{}$, respectively, without the
$f(x)$ symbol).  Let $e_m$ be a vector of zeros, except with a 1 in the $m$th
element, and let $\delta$ be a sufficiently small scalar
constant. A linear approximation to the $m$th column of the Hessian is ${\hess{m}{}\approx\left(\nabla f(x+\delta
  e_m)- \nabla f(x)\right)/\delta}$.  Estimating a dense Hessian in this way involves at least $M+1$ calculations of the
gradient: one for the gradient at $x$, and one after perturbing each
of the $M$ elements of $x$, one at a time.\footnote{More accurate
  approximations using finite differences require more gradient evaluations.  We will consider
  only forward finite differences. In Section~\ref{sec:complex} we
  introduce the complex step method that, under certain conditions,
  can generate more accurate estimates with the same number of
  gradient evaluations using forward finite differences.} In a sparse matrix, most of the
elements are constrained to zero.  Depending
on the sparsity pattern of the Hessian, those constraints may let us recover
the Hessian with fewer gradient evaluations by perturbing multiple
elements of $x$ together.  For some sparsity patterns, estimating a Hessian in
this way can be profoundly efficient.  In fact, for the hierarchical models that we consider in
this paper, the number of gradient evaluations does not increase with
the number of additional heterogeneous units.

@@

This article proceeds as follows.  First, we present
some background information about numerical differentiation, and
sparse matrices in \proglang{R}, in Section~\ref{sec:background}.  In
Section~\ref{sec:using}, we explain how to use the package.
Section~\ref{sec:algorithms} explains the underlying algorithms, and
Section~\ref{sec:timing} demonstrates the scalability of those algorithms.

@@

The \pkg{sparseHessianFD} algorithms assign variables to groups
before computing the values of the Hessian.  This is why the sparsity
pattern needs to be provided in advance.  If a non-zero element is
omitted from the sparsity pattern, the resulting estimate of the
Hessian will be incorrect.  The only problems with erroneously including a zero
element in the sparsity pattern are a possible lack of efficiency
(e.g., an increase in the number of gradient evaluations), and that
the estimated value might be close to, but not exactly, zero. The
algorithms for assigning decision variables to groups, and for
extracting nonzero Hessian elements via substitution, are described in
Section~\ref{sec:algorithms}.

@@

From the log posterior density in Equation~\ref{eq:LPD}, we can see
that $\parD{2}{\beta_i, \beta_i}\neq 0$ (one element of $\beta_i$ could be
correlated with another element of $\beta_i$), and that, for all $i$,
$\parD{2}{\beta_i, \mu}\neq 0$ (because $\mu$ is the prior mean of each
$\beta_i$).  However, since the $\beta_i$ and $\beta_j$ are independently distributed, and the $y_i$ are
conditionally independent, the cross-partial derivatives
$\parD{2}{\beta_i, \beta_j}=0$ for all $i\neq j$. When $N$ is much
greater than $k$, there will be many more zero cross-partial
derivatives than non-zero.  Regardless of how each $\parD{2}{}$ is
mapped to $\hess{}{}$ through the indexing function, the Hessian of the log
posterior density will be sparse.

@@

{\setlength{\extrarowheight}{.25em}
\begin{table}[ptb]
\begin{tabularx}{\linewidth}{>{\bfseries}r X}
  \toprule
x& A numeric vector, with length $M$ at which the object will be
  initialized and tested.\\
fn,gr& \proglang{R} Functions that return the value of the
  objective function, and its gradient. The first argument is the numeric
  variable vector.  Other named arguments can be passed to \func{fn}
  and {gr} as well (see the \funcarg{...} argument below).\\
rows, cols& Sparsity pattern: integer vectors of the row and column indices of
  the non-zero elements in the \emph{lower triangle} of the Hessian.\\
direct& This argument is deprecated, and is included only for
  backwards compatibility with earlier versions.\\
delta& The perturbation amount for finite differencing of the
  gradient to compute the Hessian (the $\delta$ in Section~\ref{sec:numdiff}).  Defaults to
  \code{sqrt(.Machine\$double.eps)}.\\
index1& If \variable{TRUE} (the default), \funcarg{row} and \funcarg{col} use one-based
  indexing.  If \variable{FALSE}, zero-based indexing is used.\\
  complex&If \variable{TRUE}, the complex step method is used.
                If \variable{FALSE} (the default), a simple finite
                differencing of gradients is used.\\
...& Additional arguments to be passed to \func{fn} and \func{gr}.\\
 \bottomrule
\end{tabularx}
\caption{Arguments to the sparseHessianFD initializer.}\label{tab:init}
\end{table}
}

@@

Now we can estimate the Hessian for the
log posterior density of the model from Section~\ref{sec:example}.
For demonstration purposes, \pkg{sparseHessianFD} includes functions that compute the value
(\func{binary.f}), the gradient (\func{binary.grad}) and the Hessian
(\func{binary.hess}) of this model. We will treat the result from \func{binary.hess} as a
  ``true'' value against which we will compare the numerical
  estimates.

@@

The \pkg{sparseHessianFD} package also allows for
estimation of the Hessian using the complex step method
\citep{SquireTrapp1998} as an alternative to finite differencing. In
this method, the definition of $\Mat{Y}_c$ in Equation~\ref{eq:Yg} is replaced with
\begin{align}
  \label{eq:complexYg}
  \Mat{Y}_c&=\Imag(\nabla f(x+iG_c))
\end{align}
where $i=\sqrt{-1}$ and $\Imag$ returns the imaginary part of a complex
number. The advantage of this method is that there is no longer a
subtraction operation to generate truncation error. Thus, $\delta$ can
be made arbitrarily small, and the resulting estimate of the Hessian
is more accurate.

A call to the \class{sparseHessianFD} \method{hessian} method will
apply the complex step method if the 
\funcarg{complex=TRUE} was set in the initializer. The functions
passed as \funcarg{fn} and \funcarg{gr} must both
accept a complex argument, and return a complex
result, even though we are differentiating a real-valued function. Although \pkg{base} \proglang{R} supports complex arguments
for most basic mathematical functions, many common functions (e.g.,  \func{gamma}, \func{log1p},
\func{expm1}, and the probability distribution functions) do not have
complex implementations. Furthermore, the complex step method is valid
only if the function is holomorphic
(complex differentiable). The methods in \class{sparseHessianFD} do
\emph{not} check that this is the case for the function at hand. We convey the
following warning from the documentation of the \pkg{numDeriv} package
\citep{R_numDeriv}, which also implements the complex step method:
``Avoid this method if you do not know that your function is
suitable. Your mistake may not be caught and the results will be
spurious.''

Fortunately for demonstration purposes, the log posterior density in Equation~\ref{eq:LPD} is holomorphic, so we
can estimate its Hessian using the complex step method, and compute
the mean relative difference from the true Hessian.

@@

As far as we know, \pkg{numDeriv} \citep{R_numDeriv} is the only other \proglang{R} package
that computes numerical approximations to derivatives.  Like
\pkg{sparseHessianFD} it includes functions to compute Hessians from
user-supplied gradients (through the \func{jacobian} function), and
implements both the finite differencing and complex step methods.  Its
most important distinction from \pkg{sparseHessianFD} is that it treats all Hessians as dense.  Thus, we
will use  \pkg{numDeriv} as the baseline against which we can compare the performance of
\pkg{sparseHessianFD}.  All of the subsequent speed and scalability
results were generated on an Apple Mac Pro with a 2.7 GHz
12-core Intel Xeon E5 processor and 64 GB of RAM, running MacOS
version 10.12.2, and \proglang{R} version 3.3.2.

To prepare Table~\ref{tab:numDeriv}, we
estimated Hessians of the log posterior density in Equation~\ref{eq:LPD} with different numbers of heterogeneous
units ($N$) and within-unit parameters ($k$).  As in the previous
section, the total number of
variables is $M=(N+1)k$.  Table~\ref{tab:numDeriv} shows the mean
and standard deviations (across 20 replications) for the time (in
milliseconds) to compute a Hessian using each package.  The difference in run times is dramatic,
especially because the computation time for \pkg{numDeriv} grows
quadratically in the number of variables. Times
were collected using the \pkg{microbenchmark} package
\citep{R_microbenchmark}. Code to replicate Table~\ref{tab:numDeriv}
is available as an online supplement to this paper, and in the
\func{doc/} directory of the installed package.

<<echo=FALSE>>=
suppressMessages(library(xtable))
suppressMessages(library(dplyr))
load("vignette_tab4.Rdata")
@

\begin{table}[ptb]\centering
  \begin{tabularx}{1\linewidth}{@{\extracolsep{\fill} }rrr|rrrr|rrrr}
  \toprule
&&&\multicolumn{4}{c|}{\bfseries{Finite differencing}}&\multicolumn{4}{c}{\bfseries{Complex step}}\\
&&&\multicolumn{2}{c}{\pkg{numDeriv}}&\multicolumn{2}{c|}{\pkg{sparseHessianFD}}&\multicolumn{2}{c}{\pkg{numDeriv}}&\multicolumn{2}{c}{\pkg{sparseHessianFD}}\\
$N$&$k$&$M$&mean&sd&mean&sd&mean&sd&mean&sd\\
%%\midrule
<<results='asis',echo=FALSE>>=
arrange(tab4, M) %>%
xtable(digits=c(rep(0,4),rep(1,8))) %>%
    print(only.contents=TRUE, comment=FALSE, booktabs=TRUE,
      include.rownames=FALSE, include.colnames=FALSE)
@
\end{tabularx}
\caption{Computation times (milliseconds) for computing Hessians using the
  \pkg{numDeriv} and \pkg{sparseHessianFD} packages, and the finite
 difference and complex step methods, across 200 replications. Rows are ordered by the
  number of variables.}\label{tab:numDeriv}
\end{table}

@@

Computation times
for the function and gradient, as well as the setup and
partitioning times for the \class{sparseHessianFD} object, grow
linearly with the number of heterogenous units.  The time
for the Hessian grows linearly as well, and that might be partially
surprising.  We saw in Section~\ref{sec:coloring} that adding
additional heterogeneous units in a hierarchical model does not
increase the number of required gradient evaluations.  So we might
think that the time
to compute a Hessian should not increase with $N$ \emph{at all}.  The
reason it does is that each gradient evaluation takes longer.
Nevertheless, we can conclude that the
\pkg{sparseHessianFD} algorithms are quite efficient and scalable for
hierarchical models.

### 2016-12-23 nobuild/jss_revision/sparseHessianFD.tex
still working on jss revision

As far as we know, \pkg{numDeriv} \citep{R_numDeriv} is the only other \proglang{R} package
that computes numerical approximations to derivatives.  Like
\pkg{sparseHessianFD} it includes functions to compute Hessians from
user-supplied gradients (through the \func{jacobian} function), and
implements both the finite differencing and complex step methods.  Its
most important distinction from \pkg{sparseHessianFD} is that it treats all Hessians as dense.  Thus, we
will use  \pkg{numDeriv} as the baseline against which we can compare the performance of
\pkg{sparseHessianFD}.  All of the subsequent speed and scalability
results were generated on a 2013-vintage Apple Mac Pro with a 2.7 GHz
12-core Intel Xeon E5 processor and 64 GB of RAM.

To prepare Table~\ref{tab:numDeriv}, we
estimated Hessians of the log posterior density in Equation~\ref{eq:LPD} with different numbers of heterogeneous
units ($N$) and within-unit parameters ($k$).  As in the previous
section, the total number of
variables is $M=(N+1)k$.  Table~\ref{tab:numDeriv} shows the mean
and standard deviations (across 20 replications) for the time (in
milliseconds) to compute a Hessian using each package.  The difference in run times is dramatic,
especially because the computation time for \pkg{numDeriv} grows
quadratically in the number of variables. Times
were collected using the \pkg{microbenchmark} package
\citep{R_microbenchmark}. Code to replicate Table~\ref{tab:numDeriv}
is in the file \filename{replication.R}, which is available as an
online supplement to this paper.

\begin{Schunk}
\begin{Sinput}
R> library(xtable)
R> library(dplyr)
R> load("vignette_tab4.Rdata")
\end{Sinput}
\end{Schunk}

@@

Computation times
for the function and gradient, as well as the setup and
partitioning times for the \class{sparseHessianFD} object, grow
linearly with the number of heterogenous units.  The time
for the Hessian grows linearly as well, and that might be partially
surprising.  We saw in Section \ref{sec:coloring} that adding
additional heterogeneous units in a hierarchical model does not
increase the number of required gradient evaluations.  So we might
think that the time
to compute a Hessian should not increase with $N$ \emph{at all}.  The
reason it does is that each gradient evaluation takes longer.
Nevertheless, we can conclude that the
\pkg{sparseHessianFD} algorithms are quite efficient and scalable for
hierarchical models.

>>>>

As far as we know, \pkg{numDeriv} \citep{R_numDeriv} is the only other \proglang{R} package
that computes numerical approximations to derivatives.  Like
\pkg{sparseHessianFD} it includes functions to compute Hessians from
user-supplied gradients (through the \func{jacobian} function), and
implements both the finite differencing and complex step methods.  Its
most important distinction from \pkg{sparseHessianFD} is that it treats all Hessians as dense.  Thus, we
will use  \pkg{numDeriv} as the baseline against which we can compare the performance of
\pkg{sparseHessianFD}.  All of the subsequent speed and scalability
results were generated on an Apple Mac Pro with a 2.7 GHz
12-core Intel Xeon E5 processor and 64 GB of RAM, running MacOS
version 10.12.2, and \proglang{R} version 3.3.2.

To prepare Table~\ref{tab:numDeriv}, we
estimated Hessians of the log posterior density in Equation~\ref{eq:LPD} with different numbers of heterogeneous
units ($N$) and within-unit parameters ($k$).  As in the previous
section, the total number of
variables is $M=(N+1)k$.  Table~\ref{tab:numDeriv} shows the mean
and standard deviations (across 20 replications) for the time (in
milliseconds) to compute a Hessian using each package.  The difference in run times is dramatic,
especially because the computation time for \pkg{numDeriv} grows
quadratically in the number of variables. Times
were collected using the \pkg{microbenchmark} package
\citep{R_microbenchmark}. Code to replicate Table~\ref{tab:numDeriv}
is available as an online supplement to this paper, and in the
\func{doc/} directory of the installed package.

@@

Computation times
for the function and gradient, as well as the setup and
partitioning times for the \class{sparseHessianFD} object, grow
linearly with the number of heterogenous units.  The time
for the Hessian grows linearly as well, and that might be partially
surprising.  We saw in Section~\ref{sec:coloring} that adding
additional heterogeneous units in a hierarchical model does not
increase the number of required gradient evaluations.  So we might
think that the time
to compute a Hessian should not increase with $N$ \emph{at all}.  The
reason it does is that each gradient evaluation takes longer.
Nevertheless, we can conclude that the
\pkg{sparseHessianFD} algorithms are quite efficient and scalable for
hierarchical models.
